---
title: "Solution and notes for the assignment Week 3"
author: "Torbjørn Ergon"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment Task

> Fit models to predict body height from a person's sex and neck circumference. Plot the predictions and write some text to interpret what you see. What do you conclude about the relationship between neck circumference and body height, and how is this pattern different from the relationship between foot length and height?


## Reading in the data

```{r}
student_means = read.csv("student_means.csv")
```

## Looking at the data

Before fitting any models to the data it's a good idea to look at a scatter plot (with the response variable on the y-axis, the continuous predictor variable (`Foot`) on the x-axis, and the categorical predictor variable (`Sex`) shown as colours (red is 'female', blue is 'male'):

```{r}
plot(student_means$Neck_circumference,student_means$Height,col=ifelse(student_means$Sex=="female", "red", "blue"))
```

We now fit a series of models with `Sex` and `Neck_circumference` as predictor variable. For each model we will look at the summary of the model and the confidence intervals of the parameters, as well as the "Multiple R-squared" value, to interpret the results.

### Only sex as a predictor variable (this was also fitted in the tutorial)

```{r}
fit_Sex = lm(Height ~ Sex, data = student_means)
summary(fit_Sex)
confint(fit_Sex)
```

Mean height of males is 15.8 cm higher than mean height of females. The 95% confidence interval for this difference is 12.6 cm to 19.0 cm. This confidence interval refers the *statistical population* that the sample of students represent. Since we have not done any random sampling, it is not clear what this statistical population is, but it would not be unreasonable to assume that these students are representative for all students at the University of Oslo - or at least each sex group is (it would be much more questionable to assume that the results are representative for the Norwegian population - why?). The overall mean height in the sample (and the confidence interval for this mean) would *not* be representative for all students at the University of Oslo because there are a higher proportion of females in the course than at the University as a whole.

The multiple R-squared value tells us that sex accounts for 71% of the variance in the height measurements in the data (i.e., the residual variance when accounting for the difference in mean height of the sexes is 29% of the variance in the height measurements).

### Only neck_circumference as a predictor variable

```{r}
fit_Neck = lm(Height ~ Neck_circumference, data = student_means)
summary(fit_Neck)
confint(fit_Neck)
```

Increasing neck circumference with 1 cm increases mean height by 1.5 cm (95% c.i.: 0.8 cm to 2.1 cm). This model explains 34% of the variance.

### Additive model
```{r}
fit_Sex_Neck = lm(Height ~ Sex + Neck_circumference, data = student_means)
summary(fit_Sex_Neck)
confint(fit_Sex_Neck)
```

One cm increase in neck circumference within sexes (i.e., when the difference among sexes are accounted for) is estimated to *decrease* height by 0.2 cm. The 95% confidence interval for the effect of increasing neck circumference by 1 cm ranges from a 0.8 cm *decrease* to a 0.4 cm *increase*. Note also that adding neck circumference as a predictor variable to the model with only sex as a predictor variable increase the proportion of variance explained from 71.3% to only 71.5%, but the estimate for the difference between sexes becomes more uncertain (i.e., less precise); standard error increases from 1.57 cm to 2.33 cm.

To better understand what is going on here, we can make a plot of the fitted predictions with confidence intervals from the additive model. Here we also add black lines for the predictions from the model with only neck circumference as the predictor variable:

```{r}
model_fit = fit_Sex_Neck
Title = "Additive model"
x_range_F = range(student_means$Neck_circumference[student_means$Sex=="female"])
x_range_M = range(student_means$Neck_circumference[student_means$Sex=="male"])
pred_data_F = data.frame(Sex = "female", Neck_circumference = seq(x_range_F[1], x_range_F[2], length.out = 50))
pred_data_M = data.frame(Sex = "male", Neck_circumference = seq(x_range_M[1], x_range_M[2], length.out = 50))
pred_data = rbind(pred_data_F, pred_data_M)
pred = predict(model_fit, pred_data, interval = "confidence")
pred = cbind(pred_data, pred)

with(student_means,
     plot(Neck_circumference, Height, type="n", xlab="Neck_circumference (cm)", ylab = "Height (cm)")
     )
with(student_means, points(Neck_circumference, Height, col=ifelse(Sex=="female", "red", "blue")))
with(pred[pred$Sex=="female",], lines(Neck_circumference, fit, col="red"))
with(pred[pred$Sex=="female",], lines(Neck_circumference, lwr, col="red", lty=2))
with(pred[pred$Sex=="female",], lines(Neck_circumference, upr, col="red", lty=2))
with(pred[pred$Sex=="male",], lines(Neck_circumference, fit, col="blue"))
with(pred[pred$Sex=="male",], lines(Neck_circumference, lwr, col="blue", lty=2))
with(pred[pred$Sex=="male",], lines(Neck_circumference, upr, col="blue", lty=2))

# Adding the fitted predictions from the model with only neck circumference as a predictor variable
pred_data_Neck = data.frame(Neck_circumference = seq(x_range_F[1], x_range_M[2], length.out = 50))
pred_Neck = cbind(pred_data_Neck, predict(fit_Neck, pred_data_Neck, interval = "confidence"))
with(pred_Neck, lines(Neck_circumference, fit))
with(pred_Neck, lines(Neck_circumference, lwr, lty=2))
with(pred_Neck, lines(Neck_circumference, upr, lty=2))

legend("topleft", legend=c("Females", "Males", "Overall"), pch=c(1,1, NA), lty = c(NA,NA,1), col = c("red", "blue", "black"))
title(Title)
```

It is not really wrong to say that people with thicker necks tend to be taller than people with thinner necks (black line). However, this is only due to a correlation among groups (sexes in this case); males tend to be both taller and have thicker necks than females. There is no evidence of such an association withing groups (the correlation is weak, and it may well be of opposite sign). It is a common mistake in data analysis (and a common mistake by journalists, politicians and common people, as well as professional researchers) to interpret patterns in data to be due to patterns/mechanisms at the individual level, when the patterns are in fact caused by a correlation among groups instead. Such confusions are referred to as "Simpson's paradox" and "the fallacy of ecological correlations" (the latter especially in social sciences). We will give more examples of this in the lectures later in the course. This also relates to the concept of "confounding": In the model with only neck circumference, we may say that the effect of neck thickness is confounded by the effect of sex (imagine studying a species where it was difficult to sex the individuals).

Note that there is not a striking lack of fit for the "neck only" model (black line above):

```{r}
plot(fit_Neck)
```


### Interaction model

```{r}
fit_Sex_Neck_inter = lm(Height ~ Sex * Neck_circumference, data = student_means)
summary(fit_Sex_Neck_inter)
confint(fit_Sex_Neck_inter)
```

Here we see that difference in the slope for males and females has a 95% confidence interval that includes zero, so this difference is not statistically significant. However, the confidence interval for the difference between the sexes is very wide (remember that the units is in cm). Hence, we can by no means conclude that males and females have the same (or nearly the same) slope in the population! To misinterpret lack of statistical significance to mean "no difference" is a **very common mistake that is important to avoid**. We can just conclude that we do not have enough data to estimate the difference precisely enough to make any inference, which is a very different thing. When we don't have enough data to make such "fine grained" inference or predictions, we better use more simplified models.

Plot of fitted predictions with confidence intervals, interaction model:
```{r}
model_fit = fit_Sex_Neck_inter
Title = "Interaction model"
x_range_F = range(student_means$Neck_circumference[student_means$Sex=="female"])
x_range_M = range(student_means$Neck_circumference[student_means$Sex=="male"])
pred_data_F = data.frame(Sex = "female", Neck_circumference = seq(x_range_F[1], x_range_F[2], length.out = 50))
pred_data_M = data.frame(Sex = "male", Neck_circumference = seq(x_range_M[1], x_range_M[2], length.out = 50))
pred_data = rbind(pred_data_F, pred_data_M)
pred = predict(model_fit, pred_data, interval = "confidence")
pred = cbind(pred_data, pred)

with(student_means,
     plot(Neck_circumference, Height, type="n", xlab="Neck_circumference (cm)", ylab = "Height (cm)")
     )
with(student_means, points(Neck_circumference, Height, col=ifelse(Sex=="female", "red", "blue")))
with(pred[pred$Sex=="female",], lines(Neck_circumference, fit, col="red"))
with(pred[pred$Sex=="female",], lines(Neck_circumference, lwr, col="red", lty=2))
with(pred[pred$Sex=="female",], lines(Neck_circumference, upr, col="red", lty=2))
with(pred[pred$Sex=="male",], lines(Neck_circumference, fit, col="blue"))
with(pred[pred$Sex=="male",], lines(Neck_circumference, lwr, col="blue", lty=2))
with(pred[pred$Sex=="male",], lines(Neck_circumference, upr, col="blue", lty=2))
legend("topleft", legend=c("Females", "Males"), pch=c(1,1), col = c("red", "blue"))
title(Title)
```

Presenting such a figure invites misinterpretations; many people will interpret this as evidence that males have a positive slope, while females have a negative slope, but such a conclusion is unjustified (the predictions are way to uncertain to predict this) - remember that we seek to make inference about the statistical population and not just the sample.
