---
title: "Week 2 - Exercises"
output: html_document
---
<style>
div.green { background-color:lightgreen; border-radius: 5px; padding: 8px;}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercises

1. Draw 10 random values from a normal distribution with mean 12 and standard deviation 2.

<div class="green">
```{r}
(x = rnorm(n=10, 12, 2))
```
</div>

a) What do you expect the variance of the random values to be?

<div class="green">    
```{r}
2^2
```
Expected variance = 4
</div>
    
b) Compute the empirical variance of the random values. Is the variance close to what you expected? If not, why not? Repeat this with 10 new random values.

<div class="green">
```{r}
# sd(x)^2
var(x)
```
</div>

c) Compute the variance of one million values drawn from the same distribution. Is this closer to the expected value? Why?

<div class="green">    
```{r}
x = rnorm(n=1000000, 12, 2)
var(x)
```
Much closer to the expected value due to the "law of large numbers"
</div>

2. Compute the values `2^i` when `i` goes from 1 to 10 in a for-loop (you should get the same results as when running `2^(1:10)` in R).

<div class="green">
```{r}
y = rep(NA,10)
for(i in 1:10){
  y[i] = 2^i
}
y

2^(1:10)

```
</div>

3. Do the same as above, but with `i` going from 0 to 10.

<div class="green">
```{r}
y = rep(NA,11)
for(i in 0:10){
  y[i+1] = 2^i # NB! Index i+1 here as indexing starts on 1 in R
}
y

2^(0:10)

```
</div>

4. Use R to simulate the probability distribution of the sum of two independent dice throws and plot a histogram. **HINTS:**
    + You can simulate one dice throw with `sample(1:6, 1)` and two independent dice throws with `sample(1:6, 2, replace = T)` (Why do we need to use `replace = T`? Remember that you can look up the description of any R-function in the 'help' (e.g. type `?sample`))
    + The `hist()` function (using default arguments) does not produce a nice histogram with integer values. Instead, you can use `table` to count how any vector elements you have of each value (try e.g. `table(c(1,1,2,3))`), and then make a bar-plot of these counts (table values) using `barplot()`.
    
<div class="green">
```{r}
n = 1000000

sum_z = rep(n)
for(i in 1:n){
  z = sample(1:6, 2, replace = TRUE)
  sum_z[i] = sum(z) # z[1] + z[2]
}

tab = table(sum_z)   # Frequency counts
probs = tab/sum(tab) # Proportions 

hist(sum_z)

barplot(probs)

# Theoretical probabilities
# 2: 1+1
1/6*1/6
probs[1]

# 3: 1+2, 2+1
(1/6)^2 + (1/6)^2
probs[2]

# 4: 1+3, 2+2, 3+1
3*(1/6)^2
probs[3]

```
</div>